MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-02-14 Article.tex
Improved introduction, added repos urls

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions. More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and experimental techniques providing information on a bigger number of components [rnaseq, single cell, …], the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{chevalier2020synthesis,benevs2023boolean} to build large Boolean models. While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems. Furthermore, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation methods. In this effort, adapting existing software to modern HPC computing architectures is an important goal.

Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation.

>>>>

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions. More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean} to build large Boolean models. While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems. However, in order to analyze the dynamic properties of such large Boolean models, we need to efficient simulation tools. In this effort, adapting existing software to modern HPC computing architectures is an important goal.

Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation. The source codes are available at \url{https://github.com/sysbio-curie/MaBoSSG/} and \url{https://github.com/sysbio-curie/MaBoSS-env-2.0/}, together with associated benchmarking data.
